\chapter{Conclusions and Future Work}
\label{conclusions_and_future_work}

\section{Summary and Conclusions}

\subsection{Cox model}
\begin{itemize}
    \item What have I done?
    \item Why is it useful and novel?
    \item What did I find out?
    \item What are the implications?
    \item What are the limitations?
\end{itemize}

\subsection{Fusilli}

\subsection{Fusilli with MND data}

\section{Future Work}

\subsection{Illustrating Fusilli capabilities with multimodal medical data}
\begin{itemize}
    \item Applying two sources of multimodal data to fusilli show finish off its development chapter ready for a thesis chapter
    \item Either ADNI or PPMI for neurodegenerative diseases, and MIMIC CXR for critical care
    \item Halfway through data download and done some preliminary analysis
    \item Feasibility: I have access to the data, and running the fusilli pipeline is quick
\end{itemize}

\subsection{Extending work on applying Fusilli to MND prognosis prediction}
\begin{itemize}
    \item Extending the work in Chapter~\ref{fusilli_on_mnd}
    \item Doing this alongside project 1
    \item We're getting more data from collaborations and the ALS Biomarkers Study
    \item Also getting D50 measurements from collaboration with Jena, possibly allowing us to use mri from further away from diagnosis if we can estimate the alsfrsr at the scan date
    \item ALS Biomarkers study team are running analysis to get more NfL measurements from samples they still have
    \item Fine tuning parameters and architectures of the fusion models as well
    \item Motivation: fluid biomarkers are more accessible than MRI etc.
    \item Feasibility: ALS Biomarkers Study etc.
\end{itemize}

\subsection{Effect of MRI preprocessing on Fusilli prognosis prediction}
\begin{itemize}
    \item Motivation: might be better to drill down rather than using whole brain, example papers: ..
    \item Feasibility: Some methods - toolkits, etc.
\end{itemize}

\subsection{Final project options}

\subsubsection{Sensitive analysis of Fusilli prognosis prediction on varying prognosis definitions}

\subsubsection{Adding spinal cord MRI data to Fusilli prognosis prediction}
Dependent on data availability
\begin{itemize}
    \item Motivation: spinal cord is important in MND, example papers to show this
    \item Feasibility: Access to spinal cord MRI data
\end{itemize}

\subsubsection{Adding radiological report derived features to Fusilli prognosis prediction}
Dependent on data availability

\section{Timeline}
\begin{itemize}
    \item What have I done so far? Papers and conference submissions
    \item Outcomes for the rest of my PhD
    \begin{itemize}
        \item Papers
    \end{itemize}
    \item Gantt chart
\end{itemize}
